Feeding Dynamic Behavior into Optimization: Simulation-based Multi-objective Overtime Planning

### 2014-09-19 data/result/Dissertacao/Texto/Latex/Article.tex
Arrumando as pastas
Criando uma pasta para os experimentos da dissertação e separando do
artigo.

\maketitle
\begin{abstract}
IT professionals are particularly subjected to working overtime due to difficulties in estimation, requirement specification, and measuring progress during a software development life-cycle. However, excessive overtime has detrimental effects on developers and the software they produce. This contrast creates the need for overtime-planning strategies to support project managers in taking the most from as few overtime hours as possible. In this paper we introduce a formulation for the overtime-planning problem which extends the state-of-art by taking into consideration both the positive effects of overtime on productivity and its negative effects on product quality. We use heuristic optimization to explore close to optimal overtime allocations under this formulation. We present an empirical study that compared our approach with current industrial practices and a similar formulation without negative effects. Results show that suppressing the flip-side of the gains brought by overtime may lead to wrong decisions. For instance, excessive allocation of overtime in large projects may lead a manager to underestimate project cost and makespan \hl{by 5\% and 3\%, respectively}, due to longer testing activities required to fix the exceeding errors introduced by overtime. Evidence also supports an overtime-management strategy used in the industry that concentrates overtime in the second half of a project schedule to compensate for early delays - \hl{about 60\% of the overtime allocations produced under this strategy are optimal}.
\end{abstract}

@@

This paper extends the state-of-the-art by taking into account the documented impact of overtime on the quality of the services provided by developers working in extended hours. It introduces a new formulation for the overtime planning problem which combines search-based optimization with a scenario-based simulation that takes into account an increase in the number of errors injected into a software product by a team working overtime. We compared our results to current overtime management practices and provide insights to support project managers on deciding when to use overtime work. By doing this, we found empirical support for an overtime allocation practice used in the industry, that is, scientific evidence aligned with project management intuition. \hl{Our results show that, on average, 60\% of the allocations that might be built according to this strategy are optimal.} We also compared our results with a similar model that ignores the dynamics of error generation caused by overtime,  showing that project managers may take wrong decisions if the increase in error generation rates is not taken into account. For instance, \hl{excessive allocation of overtime in large projects may lead a manager to underestimate project cost and makespan, respectively, by 5\% and 3\% due to longer testing activities required to fix the exceeding errors introduced by overtime.} The primary contributions of this paper are:

@@

% Table generated by Excel2LaTeX from sheet 'tabela 4'
\begin{table}[htbp]
	\scriptsize
  \centering
  \caption{Quality indicators for each instance under our approach (NSGAII) and a model without error dynamics (NSGA$_{NE}$). Pareto fronts for NSGA$_{NE}$ cover a larger part of the objective space, leading to smaller $I_{GD}$. Solutions produced under NSGAII tend to concentrate in certain parts of the objective space}
    \begin{tabular}{llccc}
    \toprule
    \textbf{Instance} & \textbf{Config.} & \textbf{I$_{C}$} & \textbf{I$_{HV}$} & \textbf{I$_{GD}$} \\
    \midrule
    \multirow{2}[2]{*}{ACAD} 
				  & NSGAII      &         0.020 $\pm$ 0.011  & \textbf{0.530 $\pm$ 0.006} &         0.004 $\pm$ 0.001  \\
          & NSGA$_{NE}$ &         0.020 $\pm$ 0.004  &         0.333 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.00}  \\\hline
    \multirow{2}[2]{*}{WMET} 
				  & NSGAII      &         0.020 $\pm$ 0.012  & \textbf{0.544 $\pm$ 0.004} &         0.002 $\pm$ 0.001  \\
          & NSGA$_{NE}$ &         0.020 $\pm$ 0.003  &         0.334 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.000} \\\hline
    \multirow{2}[2]{*}{WAMS} 
				  & NSGAII      &         0.020 $\pm$ 0.010  & \textbf{0.498 $\pm$ 0.003} &         0.002 $\pm$ 0.001  \\
          & NSGA$_{NE}$ & \textbf{0.020 $\pm$ 0.003} &         0.337 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.000} \\\hline
    \multirow{2}[2]{*}{PSOA} 
				  & NSGAII      &         0.020 $\pm$ 0.011  & \textbf{0.471 $\pm$ 0.006} &         0.004 $\pm$ 0.001  \\
          & NSGA$_{NE}$ &         0.020 $\pm$ 0.002  &         0.338 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.000} \\\hline
    \multirow{2}[2]{*}{OMET} 
				  & NSGAII      &         0.020 $\pm$ 0.011  & \textbf{0.523 $\pm$ 0.004} &         0.002 $\pm$ 0.001  \\
          & NSGA$_{NE}$ &         0.020 $\pm$ 0.003  &         0.337 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.000} \\\hline
    \multirow{2}[2]{*}{PARM} 
				  & NSGAII      &         0.020 $\pm$ 0.018  & \textbf{0.462 $\pm$ 0.007} &         0.006 $\pm$ 0.001  \\
          & NSGA$_{NE}$ & \textbf{0.020 $\pm$ 0.003} &         0.338 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.000} \\
    \bottomrule
    \end{tabular}%
\end{table}%

@@

% Table generated by Excel2LaTeX from sheet 'tabela 6'
\begin{table}[htbp]
	\scriptsize
  \centering
  \caption{Average difference in makespan and cost for solutions involving excessive overtime for each instance. data shows that the formulation without overtime error dynamics tends to understimate both cost and makespan for most projects}
    \begin{tabular}{rccc}
    \toprule
    \multicolumn{1}{c}{\multirow{2}[2]{*}{\textbf{Instance}}} & \multirow{2}[2]{*}{\textbf{Overtime  Range}} & \multicolumn{2}{c}{\textbf{(NSGA$_{NE}>$NSGAII)}} \\

@@

Based on the analysis above, we have found evidence that optimizing overtime planning without taking into account an increase in error generation rates due to developers working overtime produces distinct results from an optimization process that considers such an increase. Next, we explore the practical implications of these findings.

@@

A first result that can be observed in the charts is the linear shape of the Pareto front calculated for NSGA$_{NE}$. This front is comprised of a set of points distributed along the main diagonal of the hypercube representing the objective space for the OPP. The shape of the Pareto front implies that there exist linear relationships between the number of overtime hours invested on the project, project makespan and cost if the dynamics in error generation rates is not taken into account. These linear relationships are incompatible with previous findings in Software Engineering research, which postulate an exponential relationship between cost (effort) and duration for non-trivial software projects \cite{Boehm:1981}. They may also make the optimization easier for the genetic algorithm, resulting in a larger pool of non-dominated solutions and probing parts of the objective space which could not be explored by the more complex search accounting for error dynamics.

@@

One might question whether we need heuristic search to answer these questions, that is, can't we find their answers relying solely on simulation? We argue that this is not possible because we are interested in relationships between different results from the simulation. If we were interested in relations between parameters and results, we could sample the selected parameters and calculate the distribution of the results. But to calculate the joint distribution of two or more results we would have to sample all parameters, which is equivalent to a complete search (all combinations of possible values are assigned to parameters) or random search (a limited number of combinations are randomly picked and assigned to parameters). Complete search is unfeasible for interesting problems due to the huge size of the objective space ($9^m$, $m$ being the number of activities). On the other hand, results from random search are outperformed by our approach, as shown in RQ1. Therefore, heuristic search complements simulation by finding the points in the objective space that represent (close to) optimal combinations of simulation results.

Regarding the last question presented above, we observe strong concentration of overtime allocation in coding and testing activities in Pareto fronts calculated for NSGAII. This concentration is consistent with error dynamics: an increase in the number of errors produced in requirement analysis or design will snowball through the subsequent activities, increasing the cost of testing later in the schedule. Here the extensible nature of simulation provides us with an interesting opportunity: one can break error propagation dynamics by using formal reviews or inspections to identify and correct errors earlier than while testing the software. The proposed simulation mechanism allows developing a scenario to represent the effects of such activities \cite{Madachy:1996} and combine them with error dynamics. The concentration of overtime in later activities is also consistent with the quality of the Pareto front built for the SH OMS. As can observed in Figure II, the Pareto front for SH has few points, but it is quite close to the front built for NSGAII. Therefore, SH can be used as a starting point for the decision-maker and can be enriched by merging its results with those calculated using our approach.

Finally, despite of having different shapes in the objective space, NSGA and NSGA$_{NE}$ present extremely high inverse correlation between the number of overtime hours and project makespan (close to -1.0 Spearman rank-order index). Indeed, NSGA$_{NE}$ presents extremely high correlation between all pairs of objectives. This leads to actionable results in the sense that algorithms might not be required to pursue optimal values for all three objectives, but only two of them: minimum makespan is observed in solutions having maximum overtime. In practical terms, discarding one objective reduces the cost of optimization, allowing for more generations to be considered and a more diverse number of solutions to be found on Pareto fronts produced by NSGAII.

\vspace{3px}
\noindent
\textbf{Validity Threats:} The design of the experimental study has taken into account potential sources of validity threats, including dealing with the randomness inherent to heuristic algorithms (all heuristic algorithms were executed 50 times before comparisons were made) and selecting proper statistical inference tests and effect-sizes for analysis (non-parametric tests and effect-sizes were used, since the data to be analyzed would not present Gaussian distribution). However, the study is not completely immune to validity threats.

@@

Simulation also adds validity threats of its own, including the use of a non-infinitesimal simulation step for continuous simulation (if a larger simulation step is used to fasten objective values calculation during the optimization) or rounding problems (if a smaller step is adopted). The need to find a balance between the computational effort required to produce simulation results and their precision may also be considered a validity threat. On its turn, optimization may also require more fitness evaluations as stop criteria, thus leading to a reevaluation of other parameters, such as population size.

@@

Optimization has also been combined with simulation for project planning purposes. Ferruci et al. \cite{ferrucci:2014} present a survey of search-based optimization applied to project management. Harman et al. \cite{harman:2009} present a general framework for project planning optimization that uses a simulator to calculate fitness functions according to a project plan and a set of constraints for staff allocation. Antoniol et al. \cite{Antoniol:2004} combine a genetic algorithm with a queuing simulator to optimize staff allocation and work package ordering on software projects. The simulator calculates project makespan taking into consideration uncertainties about the effort required for each activity, rework, and abandonment of work packages. Recently, Rodriguez et al. \cite{Rodriguez:2011} used System Dynamics models in conjunction with NSGAII to find ideal team size and schedule estimations for a software project.

@@

We observed that our approach performs significantly better (high effect-size) than the margarine and critical-path overtime management strategies. However, the practice that suggests allocating overtime in the second half of the project's schedule is competitive with our approach and may lead to good overtime allocations, specially in large projects. This behavior confirms the ``gut-feeling" of practitioners who tend to allocate overtime in the last activities to compensate for earlier delays. Therefore, by combining search-based optimization and project simulation we can help researchers to design experimental studies to investigate, evaluate, and provide scientific support for practices and assumptions used in the industry, converting intuition into evidence-based knowledge.

For future work, we consider creating new simulation scenarios to model other aspects of software projects, such as exhaustion due excessive overtime, learning curve, and inspections. We also intend to integrate the approach with project management tools, allowing further experimental analysis, particularly to engage humans with the tool and observe their reaction to the overtime allocations it suggests. We also believe that creating means to help visualizing the differences among overtime allocations proposed by our approach is an important aspect towards its adoption in practice.

>>>>

\maketitle
\begin{abstract}
IT professionals are frequently subjected to working overtime due to difficulties in estimation, requirement specification, and measuring progress during a software development life-cycle. However, excessive overtime has detrimental effects on developers and the software they produce. This contrast creates the need for overtime-planning strategies to support project managers in taking the most from as few overtime hours as possible. In this paper we introduce a formulation for the overtime-planning problem which extends the state-of-art by taking into consideration both the positive effects of overtime on productivity and its negative effects on product quality. We use heuristic optimization to explore close to optimal overtime allocations under this formulation. We present an empirical study that compared our approach with current industrial practices and a similar formulation without negative effects. Results show that suppressing the flip-side of the gains brought by overtime may lead to wrong decisions. For instance, excessive allocation of overtime may lead a manager to underestimate project cost and makespan by 6.9\% and 9.2\%, respectively, due to longer testing activities required to fix the exceeding errors introduced by tired developers. Evidence also supports an overtime-management strategy used in the industry that concentrates overtime in the second half of a project schedule to compensate for delays - up to 67\% of the overtime allocations produced under this strategy are optimal.
\end{abstract}

@@

This paper extends the state-of-the-art by taking into account the documented impact of overtime on the quality of the services provided by developers working in extended hours. It introduces a new formulation for the overtime planning problem which combines search-based optimization with a scenario-based simulation that takes into account an increase in the number of errors injected into a software product by a team working overtime. We compared our results to current overtime management practices and provide insights to support project managers on deciding when to use overtime work. By doing this, we found empirical support for an overtime allocation practice used in the industry, that is, scientific evidence aligned with project management intuition. Our results show that up to 67\% of the allocations that might be built according to this strategy are optimal. We also compared our results with a similar model that ignores the dynamics of error generation caused by overtime,  showing that project managers may take wrong decisions if the increase in error generation rates is not taken into account. For instance, excessive allocation of overtime may lead a manager to underestimate project cost and makespan, respectively, by 6.9\% and 9.2\% due to longer testing activities required to fix the exceeding errors introduced by overtime. The primary contributions of this paper are:

@@

% Table generated by Excel2LaTeX from sheet 'tabela 4'
\begin{table}[htbp]
	\scriptsize
  \centering
  \caption{Quality indicators for each instance under our approach (NSGAII) and a model without error dynamics (NSGA$_{NE}$). Pareto fronts for NSGA$_{NE}$ cover a larger part of the objective space, leading to smaller $I_{GD}$. Solutions produced under NSGAII tend to concentrate in certain parts of the objective space}
    \begin{tabular}{llccc}
    \toprule
    \textbf{Instance} & \textbf{Config.} & \textbf{I$_{C}$} & \textbf{I$_{HV}$} & \textbf{I$_{GD}$} \\
    \midrule
    \multirow{2}[2]{*}{ACAD} 
				  & NSGAII      &         0.020 $\pm$ 0.011  & \textbf{0.530 $\pm$ 0.006} &         0.004 $\pm$ 0.001  \\
          & NSGA$_{NE}$ &         0.020 $\pm$ 0.004  &         0.333 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.00}  \\\hline
    \multirow{2}[2]{*}{WMET} 
				  & NSGAII      &         0.020 $\pm$ 0.012  & \textbf{0.544 $\pm$ 0.004} &         0.002 $\pm$ 0.001  \\
          & NSGA$_{NE}$ &         0.020 $\pm$ 0.003  &         0.334 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.000} \\\hline
    \multirow{2}[2]{*}{WAMS} 
				  & NSGAII      &         0.020 $\pm$ 0.010  & \textbf{0.498 $\pm$ 0.003} &         0.002 $\pm$ 0.001  \\
          & NSGA$_{NE}$ & \textbf{0.020 $\pm$ 0.003} &         0.337 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.000} \\\hline
    \multirow{2}[2]{*}{PSOA} 
				  & NSGAII      &         0.020 $\pm$ 0.011  & \textbf{0.471 $\pm$ 0.006} &         0.004 $\pm$ 0.001  \\
          & NSGA$_{NE}$ &         0.020 $\pm$ 0.002  &         0.338 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.000} \\\hline
    \multirow{2}[2]{*}{OMET} 
				  & NSGAII      &         0.020 $\pm$ 0.011  & \textbf{0.523 $\pm$ 0.004} &         0.002 $\pm$ 0.001  \\
          & NSGA$_{NE}$ &         0.020 $\pm$ 0.003  &         0.337 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.000} \\\hline
    \multirow{2}[2]{*}{PARM} 
				  & NSGAII      &         0.020 $\pm$ 0.018  & \textbf{0.462 $\pm$ 0.007} &         0.006 $\pm$ 0.001  \\
          & NSGA$_{NE}$ &         0.020 $\pm$ 0.003  &         0.338 $\pm$ 0.001  & \textbf{0.001 $\pm$ 0.000} \\
    \bottomrule
    \end{tabular}%
\end{table}%

@@

% Table generated by Excel2LaTeX from sheet 'tabela 6'
\begin{table}[htbp]
	\scriptsize
  \centering
  \caption{Average difference in makespan and cost for solutions involving excessive overtime for each instance. Data shows that the formulation without overtime error dynamics underestimates both project cost and makespan}
    \begin{tabular}{rccc}
    \toprule
    \multicolumn{1}{c}{\multirow{2}[2]{*}{\textbf{Instance}}} & \multirow{2}[2]{*}{\textbf{Overtime  Range}} & \multicolumn{2}{c}{\textbf{(NSGA$_{NE}>$NSGAII)}} \\

@@

Based on the analysis above, we have found evidence that optimizing overtime planning without taking into account an increase in error generation rates due to developers working overtime produces distinct results than an optimization process that considers such an increase. Next, we explore the practical implications of these findings.

@@

A first result that can be observed in the charts is the linear shape of the Pareto front calculated for NSGA$_{NE}$. This front is comprised of a set of points distributed along the main diagonal of the hypercube representing the objective space for the OPP. The shape of the Pareto front implies that there exist linear relationships between the number of overtime hours invested on the project, project makespan and cost if the dynamics in error generation rates is not taken into account. These linear relationships are incompatible with previous findings in Software Engineering research, which postulate an exponential relationship between cost (effort) and duration for non-trivial software projects \cite{Boehm:1981}. They may also make the optimization easier for the genetic algorithm, resulting in a larger pool of non-dominated solutions that probe parts of the objective space which could not be explored by the more complex search accounting for error dynamics.

@@

One might question whether we need heuristic search to answer these questions, that is, can't we find their answers relying solely on simulation? We argue that this is not possible because we are interested in relationships between different results from the simulation. If we were interested in relations between parameters and results, we could sample the selected parameters and calculate the distribution of the results. But to calculate the joint distribution of two or more results we would have to sample all parameters, which is equivalent to a complete search (all combinations of possible values are assigned to parameters) or random search (a limited number of combinations are randomly picked and assigned to parameters). Complete search is unfeasible for interesting problems due to the huge size of the objective space ($9^m$, $m$ being the number of activities). On the other hand, results from random search are outperformed by our approach, as shown in RQ1. Therefore, heuristic search complements simulation by finding the points in the objective space that represent (close to) optimal combinations of selected simulation results.

Regarding the last question presented above, we observe strong concentration of overtime allocation in coding and testing activities in Pareto fronts calculated for NSGAII. This concentration is consistent with error dynamics: an increase in the number of errors produced in requirement analysis or design will snowball through the subsequent activities, increasing the cost of testing later in the schedule. Here the extensible nature of simulation provides us with an interesting opportunity: one can break error propagation dynamics by using formal reviews or inspections to identify and correct errors earlier than while testing the software. The proposed simulation mechanism allows developing a scenario to represent the effects of such activities \cite{Madachy:1996} and combine it with error dynamics. The concentration of overtime in later activities is also consistent with the quality of the Pareto front built for the SH OMS. As can observed in Figure II, the Pareto front for SH has few points, but it is quite close to the front built for NSGAII. Therefore, SH can be used as a starting point for the decision-maker and can be enriched by merging its results with those calculated using our approach.

Finally, despite of having different shapes in the objective space, NSGA and NSGA$_{NE}$ present extremely high inverse correlation between the number of overtime hours and project makespan (close to -1.0 Spearman rank-order index). Indeed, NSGA$_{NE}$ presents extremely high correlation between all pairs of objectives. This leads to actionable results in the sense that algorithms might not be required to pursue optimal values for all three objectives, but only two of them: minimum makespan is observed in solutions having maximum overtime. In practical terms, discarding one objective reduces the cost of optimization, allowing for more generations to be considered in the same time frame and a more diverse number of solutions to be found on Pareto fronts produced by NSGAII.

\vspace{3px}
\noindent
\textbf{Validity Threats:} The design of the experimental study has taken into account potential sources of validity threats, including dealing with the randomness inherent to heuristic algorithms (all heuristic algorithms were executed 50 times before comparisons were made) and selecting proper statistical inference tests and effect-sizes for analysis (non-parametric tests and effect-sizes were used, since the data to be analyzed did not present Gaussian distribution). However, the study is not completely immune to validity threats.

@@

Simulation also adds validity threats, such as the use of a non-infinitesimal simulation step for continuous simulation (if a larger simulation step is used to fasten objective values calculation) or rounding problems (if a smaller step is adopted). Thus, the need to find a balance between the computational effort required to produce simulation results and their precision may be considered a validity threat. Optimization, on its turn, may require more fitness evaluations than we have selected as stop criteria, leading to a revaluation of other parameters, such as population size.

@@

Optimization has also been combined with simulation for project planning purposes. Ferruci et al. \cite{ferrucci:2014} present a survey on search-based optimization applied to project management. Harman et al. \cite{harman:2009} present a general framework for project planning optimization that uses a simulator to calculate fitness functions according to a project plan and a set of constraints for staff allocation. Antoniol et al. \cite{Antoniol:2004} combine a genetic algorithm with a queuing simulator to optimize staff allocation and work package ordering on software projects. The simulator calculates project makespan taking into consideration uncertainties about the effort required for each activity, rework, and abandonment of work packages. Recently, Rodriguez et al. \cite{Rodriguez:2011} used System Dynamics models in conjunction with NSGAII to find ideal team size and schedule estimations for a software project.

@@

We observed that our approach performs significantly better (high effect-size) than the margarine and critical-path overtime management strategies. However, the practice that suggests allocating overtime in the second half of the project's schedule is competitive with our approach and may lead to good overtime allocations, specially for large projects. This behavior confirms the ``gut-feeling" of practitioners who tend to allocate overtime in the last activities to compensate for delays. Therefore, by combining search-based optimization and project simulation we can help researchers to design experimental studies to investigate, evaluate, and provide scientific support for practices and assumptions used in the industry, converting intuition into evidence-based knowledge.

For future work, we consider creating new simulation scenarios to model other aspects of software projects, such as exhaustion due to excessive overtime, learning curve, and inspections. We also intend to integrate the approach with project management tools, allowing further experimental analysis, such as to engaging humans with the tool and observing their reaction to the distinct overtime allocations it suggests. We also believe that creating means to help visualizing the differences among overtime allocations proposed by our approach is an important aspect towards its adoption in practice.
